---
title: "Regression Live Coding"
author: "Pepa Aran"
date: "2023-04-24"
output: html_document
---

# Session 9 - Regression

Live-coding example of a modeling workflow. 

```{r}
# Load necessary packages
library(ggplot2)
```

## Load data

Example `CO2` dataset from the R `datasets` package. 
```{r}
?CO2
CO2
```

## Exploratory data analysis

Take a look at the structure of the data: types of variables, ranges, etc.
Visualize relationships between variables, with the modelling task in mind.
We want to regress CO2 uptake against other covariates.
```{r}
summary(CO2)
```

```{r}
ggplot(data = CO2) +
  geom_point(aes(
    x = conc,
    y = uptake,
    color = Treatment,
    shape = Type
  )) +
  labs(x = "Concentration", y = "CO2 uptake")
```

## First simple model: Linear regression

Task: Regress CO2 uptake on ambient CO2 concentration.
```{r}
mod_linear <- lm(uptake ~ conc,
                 data = CO2)

summary(mod_linear)
```

```{r}
ggplot(data = CO2) +
  geom_point(aes(
    x = conc,
    y = uptake,
    color = Treatment,
    shape = Type
  )) +
  labs(x = "Concentration", y = "CO2 uptake") +
  geom_smooth(
    aes(x = conc,
        y = uptake),
    method = "lm",
    se = FALSE
  )
```

### Model diagnosis

How does the linear regression fit the data? 
```{r}
plot(mod_linear, 1) # residuals vs fitted
plot(mod_linear, 2) # QQplot
plot(mod_linear, 5) # residuals vs leverage

?plot.lm            # easy diagnostic plots for linear regression
```

## Improve model fit

Instead of a straight line, can we use another funcional form?
```{r}
curve(log)
curve(x^2, xlim = c(-2, 2))
```

Model the CO2 uptake usign a logarithmic transformation of the ambient
concentration of CO2.
```{r}
mod_log <- lm(uptake ~ log(conc), data = CO2)

summary(mod_log)
```

Plot regression and check diagnostic plots (now the residuals shouldn't show
the curved trend we saw before).
```{r}
ggplot(data = CO2) +
  geom_point(aes(
    x = conc,
    y = uptake,
    color = Treatment,
    shape = Type
  )) +
  labs(x = "Concentration", y = "CO2 uptake") +
  geom_smooth(
    aes(x = conc,
        y = uptake),
    method = "lm",
    formula = y ~ log(x),
    se = FALSE
  )

plot(lm_log, 1)
plot(lm_log, 2)
plot(lm_log, 5)
```

## Include categorical covariates

Use sensible categorical covariates as predictive variables.
Pay attention to the encoding of the parameters.
```{r}
mod_cat <- lm(uptake ~ log(conc) + Treatment + Type, data = CO2)

summary(mod_cat)
```

```{r}
ggplot(data = CO2) +
  geom_point(aes(
    x = conc,
    y = uptake,
    color = Treatment,
    shape = Type
  )) +
  labs(x = "Concentration", y = "CO2 uptake") +
  geom_smooth(
    aes(x = conc,
        y = uptake,
        color = Treatment,
        lty = Type),
    method = "lm",
    formula = y ~ log(x),
    se = FALSE
  )
# Note: This code produces the plot for the regression with interactions.
# See how the logarithmic curves are not parallel and and have different
# rates of increase? For a model without interaction terms, we would have
# parallel curves (only an intercept is fit for the categorical variables).
```

Let's finally also include interaction terms.
```{r}
mod_cat_interaction <- lm(uptake ~ log(conc) * Treatment * Type, data = CO2)

summary(mod_cat_interaction)
```

## Stepwise regression

Automatically search for the best fitting regression.
```{r}
step(
  object = lm(uptake ~ 1, data = CO2),                     # lowest bound model (empty)
  scope = uptake ~ conc * log(conc) * Type * Treatment ,   # upper bound (all interactions)
  direction = "forward"
)
```