# Regression and classification {#regressionclassification}

**Chapter lead author: Pepa Aran**

## Learning objectives

After completing this tutorial, you will be able to:

-   Understand the basics of regression and classification models
-   Fit linear and logistic regression models in R
-   Choose and calculate relevant model performance metrics
-   Evaluate and compare regression models
-   Detect data outliers
-   Select best predictive variables

Contents of this Chapter are inspired and partly adopted by the
excellent book [Hands-On Machine Learning in R by Boehmke &
Greenwell](https://bradleyboehmke.github.io/HOML/).

## Tutorial

### Types of models

Models try to explain relationships between variables through a
mathematical formulation, particularly to predict a given target
variable using other explanatory variables, also called predictors.
Generally, we say that the target variable $Y$ is a function (denoted
$f$) of a set of explanatory variables $X_1, X_2, \dots, X_p$ and some
model parameters $\beta$. Models can be represented as:
$$Y \sim f(X_1, X_2, \dots, X_p, \beta)$$

This is a very general notation and depending on the structure of these
components, we get to different modelling approaches.

The first distinction comes from the type of target variable. Whenever
$Y$ is a continuous variable, we are facing a *regression* problem. If
$Y$ is categorical, we talk about *classification*.

|                 | Regression                                                           | Classification                                  |
|----------------------|-------------------------|-------------------------|
| Target variable | Continuous                                                           | Categorical                                     |
| Common models   | Linear regression, polynomial regression, KNN, tree-based regression | Logistic regression, KNN, SVM, tree classifiers |
| Metrics         | RMSE, $R^2$, adjusted $R^2$, AIC, BIC                                | Accuracy, precision, AUC, F1                    |

### Regression

In this section, we will introduce the most basic regression model:
linear regression. We'll explain how to fit the model with R, how to
include categorical predictors and polynomial terms. Finally, several
performance metrics for regression models are presented.

#### Linear regression

**Theory**

Let's start with the simplest model: linear regression. You probably
have studied linear regression from a statistical perspective, here we
will take a data-fitting approach.

For example, we can try to explain the relationship between GPP and
short wave radiation, like in the Chapter \@ref(datavis). The figure
below shows a cloud of data points, and a straight line predicting GPP
based on observed shortwave radiation values.

```{r message=FALSE, warning=FALSE}
# read and format data from Ch 3
hhdf <- readr::read_csv("./data/FLX_CH-Lae_FLUXNET2015_FULLSET_HH_2004-2006_CLEAN.csv")

set.seed(2023)
gg1 <- hhdf |>
  sample_n(2000) |>  # to reduce the dataset
  ggplot(aes(x = SW_IN_F, y = GPP_NT_VUT_REF)) +
  geom_point(size = 0.75, alpha=0.4) +
  geom_smooth(method = "lm", color = "red") +
  labs(x = expression(paste("Shortwave radiation (W m"^-2, ")")), 
       y = expression(paste("GPP (gC m"^-2, "s"^-1, ")"))) +
  theme_classic()

segment_points <- data.frame(x0 = 332, y0 = 3.65, y_regr = 8.77)

gg1 +
  geom_segment(aes(x = x0, y = y0, xend = x0, yend = y_regr), 
               data = segment_points,
               color = "blue", lwd = 1.2, alpha = 0.8)
```

We want to find the best straight line that approximates a cloud of data
points. For this, we assume a linear relationship between a single
explanatory variable $X$ and our target $Y$: $$
Y_i \sim \beta_0 + \beta_1 X_i, \;\;\; i = 1, 2, ...n \;,
$$ where $Y_i$ is the i-th observation of the target variable, and $X_i$
is the i-th value of the (single) predictor variable. $n$ is the number
of observations we have and $\beta_0$ and $\beta_1$ are constant
coefficients (model parameters). We call $\beta_0$ the intercept and
$\beta_1$ the slope of the regression line. Generally, $\hat{Y}$ denotes
the model prediction.

Fitting a linear regression is finding the values for $\beta_0$ and
$\beta_1$ such that, on average over all points, the distance between
the line at $X_i$, that is $\beta_0 + \beta_1 X_i$ (blue segment in the
plot above), and the observed value $Y_i$, is as small as possible.
Mathematically, this is minimizing the sum of the square errors, that
is: $$ 
\min_{\beta_0, \beta_1} \sum_i (Y_i - \beta_0 - \beta_1 X_i)^2 .
$$ This linear model can be used to make predictions on new data, which
are obtained by $\hat{Y}_\text{new} = \beta_0 + \beta_1 X_\text{new}$.
When the new data comes from the same distribution as the data used to
fit the regression line, this should be a good prediction.

It's not hard to imagine that the univariate linear regression can be
generalized to a multivariate linear regression, where we assume that
the target variable is a linear combination of $p$ predictor variables:
$$Y \sim \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \; ... \; + \beta_p X_p \;.$$
Note that here, $X_1, \dots, X_p$ and $Y$ are vectors of length
corresponding to the number of observations in our data set ($n$ - as
above). Analogously, calibrating the $p+1$ coefficients
$\beta_0, \beta_1, \beta_2, ..., \beta_p$ is to minimize the sum of
square errors $\min_{\beta} \sum_i (Y_i - \hat{Y}_i)^2$.

While the regression is a line in two-dimensional space for the
univariate case, it is a plane in three-dimensional space for bi-variate
regression, and hyperplanes in higher dimensions.

**Implementation in R**

To fit a univariate linear regression model in R, we can use the `lm()`
function. Already in Chapter \@ref(datawrangling), we created linear
models by doing:

```{r message=F}
# numerical variables only, remove NA
df <- hhdf %>%
  dplyr::select(-starts_with("TIMESTAMP")) %>%
  tidyr::drop_na()

# fit univariate linear regression
linmod1 <- lm(GPP_NT_VUT_REF ~ SW_IN_F, data = df)
```

Here, `GPP_NT_VUT_REF` is $Y$, and `SW_IN_F` is $X$. We can include
multiple predictors for a multivariate regression, for example as:

```{r}
# fit multivariate linear regression
linmod2 <- lm(GPP_NT_VUT_REF ~ SW_IN_F + VPD_F + TA_F, data = df)
```

or all available features in our data set (all columns other than
`GPP_NT_VUT_REF` in `df`) as:

```{r}
linmod3 <- lm(GPP_NT_VUT_REF ~ ., data = df)
```

`linmod*` is now a model object of class `"lm"`. It is a list containing
the following components:

```{r}
ls(linmod1)
```

Enter `?lm` in the console for a complete documentation of these
components and other details of the linear model implementation.

R offers a set of generic functions that work with this type of object.
The following returns a human-readable report of the fit. Here the
*residuals* are the difference between the observed target values and
the predicted values.

```{r}
summary(linmod1)
```

We can also extract coefficients $\beta$ with

```{r}
coef(linmod1)
```

and the residual sum of squares (which we wanted to minimize) with

```{r}
sum(residuals(linmod1)^2)
```

Although `summary()` provides a nice, human-readable output, you may
find it unpractical to work with. A set of relevant statistical
quantities are returned in a tidy format using `tidy()` from the `broom`
package:

```{r}
broom::tidy(linmod1)
```

**Model advantages and concerns**

An advantage of linear regression is that the coefficients provide
information that is straight-forward to interpret. We've seen above,
that `GPP_NT_VUT_REF` increases by 0.0255 for a unit increase in
`SW_IN_F`. Of course, the units of the coefficients depend on the units
of `GPP_NT_VUT_REF` and `SW_IN_F`. This has the advantage that the data
does not need to be normalised. That is, a linear regression model with
the same predictive skills can be found, irrespective of whether
`GPP_NT_VUT_REF` is given in g C m$^{-2}$s$^{-1}$ or in kg C
m$^{-2}$s$^{-1}$.

Another advantage of linear regression is that it's much less prone to
*overfit* than other algorithms. You'll learn more about the concept of
overfitting in Chapter \@ref(supervisedmli). Not being prone to
overfitting can also be a disadvantage: linear regression models are
often not flexible enough to be effectively fit to the data. They are
also not able to capture non-linearities in the observed relationship
and, as we'll see later in this chapter, it exhibits a poorer
performance than more complex models (e.g. polynomial regression) also
on the validation data set.

A further limitation is that least squares regression requires $n>p$. In
words: the number of observations must be greater than the number of
predictors. If this is not given, one can resort to step-wise forward
regression, where predictors are sequentially added based on which
predictor adds the most additional information at each step. 

<!-- XXXexercises XXX. -->

When multiple predictors are linearly correlated, then linear regression
cannot discern individual effects and individual predictors may appear
statistically insignificant when they would be significant if covarying
predictors were not included in the model. Such instability can get
propagated to predictions. Again, stepwise regression can be used to
remedy this problem. However, when one predictor covaries with multiple
other predictors, this may not work. For many applications in Geography
and Environmental Sciences, we deal with limited numbers of predictors.
We can use our own knowledge to examine potentially problematic
covariations and make an informed pre-selection rather than throwing all
predictors we can possibly think of at our models. Such a pre-selection
can be guided by the model performance on a validation data set (more on
that below).

An alternative strategy is to use *dimension reduction* methods.
Principal Component regression reduces the data to capture only the
complementary axes along which our data varies and therefore collapses
covarying predictors into a single one that represents their common axis
of variation. Partial Least Squares regression works similarly but
modifies the principal components so that they are maximally correlated
to the target variable. You can read more on their implementation in R
[here](https://bradleyboehmke.github.io/HOML/linear-regression.html#PCR).

#### Regression on categorical variables

In the [regression within categories](#vis_regr_cat) section of Chapter \@ref(datavis), we saw that when we separate the data into sub-plots,
hidden patterns emerge. This information is very relevant for modeling,
because it can be included in our regression model. It is crucial to
spend enough time exploring the data before you start modeling, because
it helps to understand the fit and output of the model, but also to
create models that capture the relationships between variables better.

```{r}
# create month category
df_cat <- hhdf |>
  mutate(MONTH = lubridate::month(TIMESTAMP_START)) |>
  tidyr::drop_na() |>
  dplyr::select(MONTH, GPP_NT_VUT_REF, SW_IN_F)
```

So far, we have only used continuous variables as explanatory variables
in a linear regression. It is also possible to use categorical
variables. To do this in R, such variables cannot be of class `numeric`,
otherwise the `lm()` function treats them as continuous variables. For
example, although the variable `NIGHT` is categorical with values `0`
and `1`, the model `linmod3` treats it as a number. We must make sure
that categorical variables have class `character` or, even better,
`factor`.

```{r}
# fix class of categorical variables
df_cat <- df_cat |>
  mutate(MONTH = as.factor(MONTH))
```

Now we can fit the linear model again:

```{r}
linmod_cat <- lm(GPP_NT_VUT_REF ~ MONTH + SW_IN_F, data = df_cat)
summary(linmod_cat)
```

In the fit summary, you can observe that, there are `MONTH2` to
`MONTH12` parameters. `MONTH` is a factor which can take 12 different
values: `1` to `12`. `lm()` uses one of the factor level as the
reference, in this case `1`, and fits an intercept for the other
categories. The result is a set of parallel regression lines, one for
each different month.

```{r}
df_cat |>
  mutate(MONTH_NAME = lubridate::month(as.integer(MONTH), label = TRUE)) |>
  ggplot(aes(x = SW_IN_F, y = GPP_NT_VUT_REF)) +
  geom_point(alpha = 0.2) +
  geom_smooth(formula = y ~ x + 0, method = "lm", color = "red", se = FALSE) +
  labs(x = "SW", y = "GPP") +
  facet_wrap(~MONTH_NAME) +
  theme_classic()
```

In the grid image, we can observe that GPP does not increase with SW at
the same rate every month. For example, the increase in GPP is less
steep in February than in September. To model this, we should consider a
variable slope parameter for each month or category. In R, this is
implemented by including an *interaction* term `MONTH:SW_IN_F` in the
regression formula, like this:

```{r}
linmod_inter <- lm(GPP_NT_VUT_REF ~ MONTH + SW_IN_F + MONTH:SW_IN_F, data = df_cat)
# equivalently: lm(GPP_NT_VUT_REF ~ MONTH * SW_IN_F, data = df_cat)
summary(linmod_inter)
```

#### Polynomial regression

Furthermore, the relationships between variables may be non-linear. In
the previous example, we see that the increase in GPP saturates as
shortwave radiation grows, which suggests that the true relationship
could be represented by a curve. There are many regression methods that
fit this kind of relationship, like polynomial regression,
[LOESS](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/loess)
(local polynomial regression fitting), etc.

Let's fit a simple quadratic regression model, just for the month of
August. For this we use the `poly()` function which constructs
orthogonal polynomials of a given degree:

```{r}
quadmod <- lm(GPP_NT_VUT_REF ~ poly(SW_IN_F, 2), 
              data = df_cat |>
                filter(MONTH == 8))
summary(quadmod)
```

In the following plot, you can see how the model fit for GPP in August
improves as we consider higher degree polynomials:

```{r}
df_cat |>
  filter(MONTH == 8) |>
  ggplot(aes(x = SW_IN_F, y = GPP_NT_VUT_REF)) +
  geom_point(alpha = 0.4) +
  geom_smooth(formula = y ~ x, method = "lm", aes(color = "lm"), se = FALSE) +
  geom_smooth(formula = y ~ poly(x, 2), method = "lm", 
              aes(color = "poly2"), se = FALSE) +
  geom_smooth(formula = y ~ poly(x, 3), method = "lm",
              aes(color = "poly3"), se = FALSE) +
  geom_smooth(formula = y ~ poly(x, 4), method = "lm", 
              aes(color = "poly4"), se = FALSE) +
  labs(x = "SW", y = "GPP", color = "Regression") +
  theme_classic()
```

#### Metrics for regression evaluation

Metrics measure the quality of fit between predicted and observed
values, are essential to model training (where the metric defines the
loss function, see Chapter \@ref(supervisedmlii)), and inform model
evaluation. Different metrics measure different aspects of the
model-data agreement. In other words, a single metric never captures all aspects and
patterns of the model-data agreement. Therefore, a visual inspection of
the model fit is always a good start of the model evaluation.

To get an intuitive understanding of the different abilities of different metrics, compare
the scatterplots in Fig. \@ref(fig:correlationplots) and how different aspects of the model-data
agreement are measured by different metrics. The observed target values $y$ are plotted against the predicted values $\hat{y}$ from a 
regression model, and the dashed line represents the ideal scenario: our predictions matching the data perfectly. Definitions of the metrics displayed and other metrics are given below.

```{r correlationplots, echo=FALSE, fig.cap="Correlation plots between observed and fitted target values.", fig.height=10, message=FALSE, warning=FALSE}
analyse_modobs <- function(df){
  
  # fit linear regression model to get slope
  linmod <- lm(y ~ x, data = df)
  
  # construct metrics table using the 'yardstick' library
  df_metrics <- df %>%
    
    # simply the squared correlation between truth and estimate (0,1)
    yardstick::rsq(., y, x)  |> 
    
    # root mean square error
    dplyr::bind_rows(yardstick::rmse(df, y, x)) |> 
    
    # slope of fitted vs observed
    dplyr::bind_rows( 
      tibble(.metric = "slope", 
             .estimator = "standard", 
             .estimate = coef(linmod)[2])) |> 

    # bias
    dplyr::bind_rows( 
      tibble( .metric = "bias", 
              .estimator = "standard", 
              .estimate = mean(df$y - df$x))) |> 
    
    # format to keep only 2 significant digits
    dplyr::mutate(.estimate_form = format(.estimate, digits = 2))
  
  subtitle <- bquote( 
    italic(R)^2 == .(df_metrics |> 
                       dplyr::filter(.metric == "rsq") |> 
                       pull(.estimate_form)) ~~
    RMSE == .(df_metrics |> 
                dplyr::filter(.metric == "rmse") |> 
                pull(.estimate_form)) ~~
    Bias == .(df_metrics |> 
                dplyr::filter(.metric == "bias") |> 
                pull(.estimate_form)) ~~
    Slope == .(df_metrics|> 
                 dplyr::filter(.metric == "slope")|> 
                 pull(.estimate_form))
    )
  
  # points
  df |> 
    ggplot(aes(x, y)) +
    geom_point() +
    geom_abline(intercept = 0, slope = 1, linetype = "dotted") +  # 1:1 line
    theme_classic() +
    labs(subtitle = subtitle) + 
    geom_smooth(method = 'lm', color = "red", size = 0.5, se = FALSE)  # regression fit
  
}

set.seed(1982)

# uncorrelated
gg1 <- tibble(x = rnorm(100),
              y = rnorm(100)) |> 
  analyse_modobs() +
  labs(title = "Uncorrelated") +
  ylim(-3,3) + xlim(-3, 3) +
  xlab(expression(hat(y))) +
  theme(plot.title = element_text(size = 10),
        plot.subtitle = element_text(size = 8))

# perfectly correlated - zero error
gg3 <- tibble(x = rnorm(100)) |> 
  mutate(y = x) |> 
  analyse_modobs() +
  labs(title = bquote("Perfectly correlated, zero error:" ~ y == hat(y))) +
  ylim(-3,3) + xlim(-3, 3) +
  xlab(expression(hat(y))) +
  theme(plot.title = element_text(size = 10),
        plot.subtitle = element_text(size = 8))

# perfectly correlated - scaling error
gg4 <- tibble(x = rnorm(100)) |> 
  mutate(y = 0.6 * x) |> 
  analyse_modobs() +
  labs(title = bquote("Perfectly correlated, scaling error:" ~ y == ~ .(0.6) ~ hat(y))) +
  ylim(-3,3) + xlim(-3, 3) +
  xlab(expression(hat(y))) +
  theme(plot.title = element_text(size = 10),
        plot.subtitle = element_text(size = 8))

# perfectly correlated - constant error
gg5 <- tibble(x = rnorm(100)) |> 
  mutate(y = x + 1) |>
  analyse_modobs() +
  labs(title = bquote("Perfectly correlated, constant error:" ~ y == hat(y) + 1)) +
  ylim(-3,3) + xlim(-3, 3) +
  xlab(expression(hat(y))) +
  theme(plot.title = element_text(size = 10),
        plot.subtitle = element_text(size = 8))

# correlated
gg6 <- tibble(x = rnorm(100)) |> 
  mutate(y = x + rnorm(100)) |> 
  analyse_modobs() +
  labs(title = bquote("Correlated:" ~ y == hat(y) + epsilon[0])) +
  ylim(-3,3) + xlim(-3, 3) +
  xlab(expression(hat(y))) +
  theme(plot.title = element_text(size = 10),
        plot.subtitle = element_text(size = 8))

# better correlation
gg7 <- tibble(x = rnorm(100)) |> 
  mutate(y = x + 0.7 * rnorm(100)) |> 
  analyse_modobs() +
  labs(title = bquote("Better correlated:" ~ y == hat(y) + ~ .(0.7) ~ epsilon[0])) +
  xlab(expression(hat(y))) +
  # labs(title = "Better correlated (y = x + 0.7 rnorm(100))") +
  ylim(-3,3) + xlim(-3, 3) +
  theme(plot.title = element_text(size = 10),
        plot.subtitle = element_text(size = 8))

cowplot::plot_grid(gg1, gg3, gg4, gg5, gg6, gg7, 
                   ncol = 2)
```


Common metrics used for evaluating regression fits are:

**MSE**

The mean squared error is defined, as its name suggests,
as: $$
\text{MSE} = \frac{1}{n} \sum_{i=1}^n (Y_i - \hat{Y_i})^2
$$It measures the magnitude of the errors, and is minimized to fit a
linear regression or, as we will see in Chapter
\@ref(supervisedmli), during model training when used as a loss
function. Note that since it scales with the square of the errors,
the MSE is sensitive to large errors in single points, including
outliers.

**RMSE**

The root mean squared error is, as its name suggests, the
root of the MSE: $$
\text{RMSE} = \sqrt{\text{MSE}} = \sqrt{\frac{1}{n} \sum_{i=1}^n (Y_i - \hat{Y_i})^2}
$$Like the MSE, the RMSE also measures the magnitude of
the errors and is minimized during model training. By taking the
square root of mean square errors, the RMSE is in the same units as
the data $Y$ and is less sensitive to outliers than the MSE.

**MAE**

The mean absolute error is similarly defined: $$
\text{MAE} = \frac{1}{n} \sum_{i = 1}^{n} |Y_i - \hat{Y_i}|
$$
Measuring the discrepancies between predictions and observations using absolute errors, 
instead of squared errors, gives less importance to errors of large magnitude and more importance
to small errors than the MSE would. Hence, this measures is more stable in the presence of 
outliers.

**$R^2$ - coefficient of determination**

describes the proportion of variation in $y$ that is captured by modelled values
$\hat{y_i}$. It tells us how much better our fitted values $\hat{y_i}$ are than just
taking the average of the target $\bar{y}$ as predictions. In this case, the goal is to maximize the metric, thus
trying the explain as much variation as possible. In contrast to the
MSE and RMSE, $R^2$ measures *consistency*, or *correlation*, or
*goodness of fit*. It is defined as:
$$
R^2 = 1 - \frac{\sum_i (\hat{Y}_i - Y_i)^2}{\sum_i (Y_i - \bar{Y})^2}\\
$$
When the regression model is fitted by minimizing the MSE, the $R^2$ takes values
between 0 and 1. A perfect fit is quantified by $R^2 = 1$. There is no generally
valid threshold of $R^2$ for a model to be considered "good". It
depends on the application and the nature of the data and the
data-generating process. Note that the above equation can also be
written as $R^2 = 1 - \text{MSE}/var(Y)$.

$r$ - **Pearson's correlation**

The linear association between two variables (here $y$ and $\hat{y}$) is measured by the *Pearson's correlation coefficient* $r$. 
$$
r = \frac{\sum_i (Y_i - \bar{Y}) (\hat{Y_i} - \bar{\hat{Y}}) }{\sqrt{ \sum_i(Y_i-\bar{Y})^2 \; (\hat{Y_i}-\bar{\hat{Y}})^2 } }
$$

The correlation calculated between the target $Y$ and a predictor $X$ can tell us about the predictive 
power of $X$ in a linear regression model (the higher the correlation, the more powerful). We can also compute
the correlation between the target $Y$ and the predicted values $\hat{Y}$ by a model (multivariate, or even 
not linear) to assess the adequacy of the model chosen. See Figure \@ref(fig:correlationplots) as an example. It is
noteworthy to mention that correlation is location and scale invariant, hence it will not detect 
model deviations like the ones in the middle row plots.

The squared value of the Pearson's *r* is often reported as "$R^2$" but doesn't
correspond to the definition of the coefficient of determination
given above. However, the square of the Pearson's *r* is closely
related to the coefficient of determination $R^2$. For a
linear regression, fitted minimizing the MSE, they are identical (see proof
[here](https://statproofbook.github.io/P/slr-rsq)). In subsequent chapters, we 
will use "$R^2$" to refer to the square of the Pearson's *r*
between the observed $Y$ and predicted $\hat{Y}$ values.

> Note the implementations in R.The $R^2$ reported by the generic
`summary()` function corresponds to the base-R function `cor()^2` , to
`yardstick::rsq()`, and to the definition of the square of the
*Pearson's* $r$ given above. The `yardstick::rsq_trad()` returns the coefficient of determination as
traditionally defined and is not equal to the values above, unless computed on the predicted values $\hat{y}$. 

> Sometimes the Person's version is computed between $y$ and $x$, and it leads to the 
same number due to its "location and scale invariant" property. Nevertheless, this is 
conceptually wrong, as we should look at the predictions, not the predictors: We are 
not predicting $y$ by just giving the values of $x$ instead. Hence, especially when 
using {yardstick} functions, make sure you compute the values on $\hat{y}$. When we 
have several predictors, it's already clear that we should compare $y$ to $\hat{y}$
instead of $y$ to each predictor separately.

```{r}
# generate correlated random data 
set.seed(1982)
df <- tibble(x = rnorm(100)) |> 
  mutate(y = x + rnorm(100)) |>
  mutate(y_fitted = lm(y ~ x)$fitted.values)

# implementations using Pearson's correlation
summary(lm(y ~ x, data = df))$r.squared
cor(df$y, df$x)^2 # remember: location and scale invariant
yardstick::rsq(df, y, x) |> pull(.estimate)
(sum((df$x - mean(df$x))*(df$y - mean(df$y))))^2/
  (sum((df$y - mean(df$y))^2)*sum((df$x - mean(df$x))^2))

# implementations using coefficient of determination definition
1 - sum((df$x - df$y)^2) / sum((df$y - mean(df$y))^2) # should be \hat{y}, not x
yardstick::rsq_trad(df, y, x) %>% pull(.estimate) # incorrect
yardstick::rsq_trad(df, y, y_fitted) %>% pull(.estimate) # correct
```

> An "$R^2$" is commonly reported when evaluating the *agreement* between observed
and predicted values of a given model. When the *correlation*
between two different variables in a sample is quantified, $r$ is
commonly used to reflect also whether the correlation is positive or
negative ($r$ can attain positive or negative values in the interval $[-1, 1]$). The
coefficient of determination can return negative values when
comparing observed and predicted values for uninformative estimates (worse than just using the average of $Y$)
and is thus not actually bound between 0 and 1. Therefore, be careful with the interpreration of "$R^2$" and
think on which variables it was computed and with which method.

**Bias**

The bias is simply the mean error:
$$
\text{Bias} = \frac{1}{n} \sum_i^n{(\hat{Y}_i - Y_i)}
$$


**Slope**

The slope refers to the slope of the linear regression line between predicted and observed values. It is returned as the second element of the vector returned by `coef(lm(..))`:
```{r}
coef(lm(y ~ y_fitted, data = df))[2]
```

<!-- XXX Exercise: In supervised machine learning, a certain metric is chosen to guide the model training (fitting). Which metric would you chose as a target for optimisation. Discuss why. XXX -->

#### Metrics for regression model comparison

In general, the aim of regression modelling is to find a model that best explains
the data - but not the random errors in the data. More complex models tend to *overfit* more than simpler models. The implication of overfitting is that the model fits the data used for model fitting well, but doesn't fit well when evaluating the predictions of the same model to new data (data not used for model fitting). In such a case, the model's *generalisability* is poor. We'll learn more about overfitting and generalisability in the context of supervised machine learning in later chapters. Often, simpler models generalise better than more complex model. The challenge is to strike a balance between complexity and generalisability. But how to find the "sweet spot" of this trade-off?

In this context it should be noted that the $R^2$ *always* increases when predictors are added to a model. Therefore, the $R^2$ is not a suitable metric for comparing models that differ with respect to their number of predictors - a factor controlling model complexity. *Cross-validation* can be regarded as the "gold-standard" for measuring model generalisability if the data is plentiful. It will be introduced in the context of supervised machine learning in Chapter \@ref(supervisedmlii). However, when the data size is small, cross validation estimates may not be robust. Without
resorting to cross validation, the effect of spuriously improving the
evaluation metric by adding uninformative predictors can also be
mitigated by penalizing the number of predictors $p$. Different metrics
are available:

**Adjusted** $R^2$

The adjusted $R^2$ discounts values of $R^2$ by the
number of predictors. It is defined as $$
{R}^2_{adj} = 1 - (1-R^2) \; \frac{n-1}{n-p-1} \;,
$$ where $n$ (as before) is the number of observations, $p$ the
number of parameters and $R^2$ the usual coefficient of
determination. Same as for $R^2$, the goal is to maximize
$R^2_{adj}$.

**AIC**

The Akaike's Information Criterion is defined in terms of
log-likelihood (covered in Quantitative Methoden) but for linear
regression it can be written as: $$
\text{AIC} = n \log \Big(\frac{\text{SSE}}{n}\Big) + 2(p+2)
$$ where $n$ is the number of observations used for estimation, $p$
is the number of explanatory variables in the model and SSE is the
sum of squared errors (SSE$= \sum_i (Y_i-\hat{Y_i})^2$). Also in
this case we have to minimize it and the model with the minimum
value of the AIC is often the best model for forecasting. Since it
penalizes having many parameters, it will favor less complex models.

**AIC**$_c$

For small values of $n$ the AIC tends to select too
many predictors. A bias-corrected version of the AIC is defined as:
$$
\text{AIC}_c = \text{AIC} + \frac{2(p + 2)(p + 3)}{n-p-3}
$$ Also AIC$_c$ is minimized for an optimal predictive model.

**BIC**

The Schwarz's Bayesian Information Criterion is defined as
$$
\text{BIC} = n \log \Big(\frac{\text{SSE}}{n}\Big) + (p+2)  \log(n)
$$ Also for BIC, the goal is to minimize it. This metric has the
feature that if there is a true underlying model, the BIC will
select that model given enough data. The BIC tends to select a model
with fewer predictors than AIC.

**Implementation in R**

Let's calculate the metrics introduced above for a few of the fitted
regression models. Some of these metrics, like $R^2$ and $R^2_{adj}$ are
given by the `summary()` function. Alternatively, the {yardstick}
package provides implementations for a few of these metrics, which we
compute below:

```{r}
compute_regr_metrics <- function(mod){
  
  p <- length(mod$coefficients)
  n <- length(mod$residuals)
  
  tibble(
    mse = mean(mod$residuals^2),
    R2 = summary(mod)$r.squared,
    R2_adj = summary(mod)$adj.r.squared,
    AIC = extractAIC(mod)[2],
    AIC_adj = extractAIC(mod)[2] + 2*(p+2)*(p+3)/(n-p-3),
    BIC = BIC(mod) # this implementation is based on log-likelihood
  )
}

list_metrics <- purrr::map(
    list(linmod1, linmod2, linmod_cat, quadmod), 
    ~compute_regr_metrics(.))
names(list_metrics) <- c("Linear model", 
                         "Linear model 2", 
                         "Linear + categories",
                         "Quadratic model")
bind_rows(list_metrics, .id = "type")
```

### Classification

Classification models predict a categorical target variable. Note that
predictors may be continuous. We will introduce a classification problem
with a binary target, since it's straightforward to generalize to
categorical variables with more than two classes. As an example, let's
observe the `CO2` dataset from the [{datasets}
package](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/zCO2.html),
which contains data from an experiment on the cold tolerance of a grass
species. We will try to classify the origin of each plant (categorical
variable `Type` with values `Quebec` or `Mississippi`) depending on the
carbon dioxide uptake rate of the plant (continuous variable `uptake`
measured in $\mu$mol m$^{-2}$s$^{-1}$). More information on the dataset can be obtained by typing `?datasets::CO2` in the console.

```{r}
gg2 <- datasets::CO2 |>
  ggplot(aes(x = uptake, y = Type, color = Type)) +
  geom_point(size = 0.75, alpha = 0.8) +
  theme_classic() +
  theme(legend.position = "none")
gg2
```

At first sight, it's easy to see that the carbon uptake is lower for the
Mississippi type. Note that other predictors can be included in the
model, but we'll focus on a single predictor. Using this example, we'll
cover logistic regression, its implementation in R and metrics for
classification.

#### Logistic regression

**Theory**

A classification problem is a bit more difficult to write mathematically
than a regression problem. Before, the mathematical representation of
`GPP_NT_VUT_REF ~ SW_IN_F` was
`GPP_NT_VUT_REF`$\;=\; \beta_0 + \beta_1$`SW_IN_F`. With the
classification model `Type ~ uptake`, we cannot just write
`Type`$\;=\; \beta_0 + \beta_1$`uptake` because `Type` is not a number.
Hence, the categorical variable must be encoded, in this case 0
represents `Quebec` and 1 represents `Mississippi`.

The next issue is that a linear model makes continuous predictions in
the entire real numbers space $(-\inf, \inf)$, but we want the
predictions to be either 0 or 1. We can transform these values to be in
the interval $[0,1]$ with a *link* function. For a binary response, it's
common to use a *logit* link function:
$$\text{logit}(z) = \frac{\exp(z)}{1+\exp(z)}.$$

```{r}
curve(exp(x)/(1+exp(x)), -5, 5, ylab = "logit(x)")
```

Combining a linear model (with any type of predictors, like for
regression) and a logit link function, we arrive to a **logistic
regression** model:
$$f(X, \beta) = \text{logit}(\beta_0 + \beta_1 X_1 + ... + \beta_p X_p) = \frac{\exp(\beta_0 + \beta_1 X_1 + ... + \beta_p X_p)}{1 + \exp(\beta_0 + \beta_1 X_1 + ... + \beta_p X_p)}.$$
This predicted value can be understood as the probability of belonging
to class 1 (in our example, `Mississippi`). A classification rule is
defined such that an observation $X_{new}$ with a predicted probability
of belonging to class 1 higher than a given *threshold* $\tau$ (i.e.
$f(X_{new}, \beta) > \tau$) will be classified as 1; and if the
predicted probability is smaller than the threshold, it will be
classified as 0.

A logistic regression model results in a linear classification rule.
This means that the $p$-dimensional space will be divided in two by a
hyperplane, and the points falling in each side of the hyperplane will
be classified as 1 or 0. In the example above with carbon uptake as
predictor, the classification boundary would be a point dividing the
real line. If we include a second predictor, we would obtain a line
diviting the 2-dimensional plane in two.

Furthermore, to fit a logistic regression model means to calculate the
maximum likelihood estimator of $\beta$ with an iterative algorithm. We
will learn more about iterative model fitting, i.e. parameter
optimization, in the context of supervised machine learning (Chapter
\@ref(supervisedmlii)).

**Implementation in R**

First, let's see how the target variable is encoded. R directly loads
the dataframe with `Type` as a factor and `Quebec` as the reference
level. R factors work such that each factor level (here `Quebec` and
`Mississippi`) corresponds to an integer value (its position given by
`levels()`, here `1` and `2` respectively). We can fit a logistic model
in R with this encoding.

```{r}
str(datasets::CO2)
levels(datasets::CO2$Type)
```

To fit a logistic regression in R we can use the `glm()` function, which
fits a generalized linear model, indicating that our target variable is
binary and the link function is a logit function. Let's see the model
output:

```{r}
logmod <- glm(Type ~ uptake,
              family = binomial(link = logit),
              data = datasets::CO2)
summary(logmod)
```

This fitted model results in a linear classification boundary
(discontinued line) that splits the predictor variables space in two.
Where that line falls depends on the choice of threshold, in this case
$\tau=0.5$ (see where the grey logistic regression line meets the dashed
threshold line). You can see it plotted below:

```{r}
beta <- coef(logmod)
# reuse previous plot with classification line
gg2.1 <- datasets::CO2 |>
  ggplot(aes(x = uptake, y = as.numeric(Type)-1, color = Type)) +
  geom_point(size = 0.75, alpha = 0.8) +
  ylab("") +
  theme_classic() +
  stat_smooth(method = "glm", color = "grey", se = FALSE,
              method.args = list(family=binomial)) +
  geom_vline(xintercept = -beta[1]/beta[2], lty = 2)

gg2.1
```

Most blue points fall to one side of the dashed classification line and
most red points to the other side; this is what we wanted. The points
that are in the wrong side of the line are misclassified by the logistic
regression model, we're trying to minimize that.

Note that, just like for linear regression, a logistic regression model
allows to use categorical explanatory variables and polynomial
transformations of the predictors to achieve better-fitting
classification models.

**Model advantages and concerns**

One advantage of logistic regression is simplicity. It's part of the
*generalized linear regression* family of models and the concept of a
link function used to build such a model can also be used for various
types of response variables (not only binary, but also count data...).
You can find more details in this [Wikipedia
article](https://en.wikipedia.org/wiki/Generalized_linear_model).

Furthermore, logistic regression allows for an interesting
interpretation of its model parameters: *odds* and *log-odds*. Odds
represent how much likely it is to find one class versus the other (e.g.
class 1 is twice as likely as class 0 whenever we have probabilities
$66\%$ vs $33\%$). The *odds* are defined as the probability of $Y$
belonging to class 1 divided by the probabiity of belonging to class 0,
and relates to the model parameters as
$$\frac{P(Y_i=1)}{P(Y_i=0)} = \exp(\beta_0+\beta_1 X_i).$$ So the
log-odds are
$$\log\left(\frac{P(Y_i=1)}{P(Y_i=0)}\right) = \beta_0+\beta_1 X_i.$$
Increases in the values of the predictors affect the odds
multiplicatively and the log-odds linearly.

It is easy to extend a logistic regression model to more than two
classes by fitting models iteratively. For example, first you classify
class 1 against classes 2 and 3; then another logistic regression
classifies class 2 against 3.

Nevertheless, logistic regression relies on statistical assumptions to
fit the parameters and interpret the fitted parameters. So whenever
these assumptions are not met, one must be careful with the conclusions
drawn. Other machine learning methods, that will be covered in Chapters
\@ref(supervisedmli) and \@ref(supervisedmlii), can also be used for
classification tasks. These offer more flexibility than logistic
regression (are not necessarily linear) and don't need to satisfy strict
statistical assumptions.

#### Metrics for classification

Measuring the quality of a classification model is based on counting how
many observations were correctly classified, rather than the distance
between the values predicted by a regression and the true observed
values. These can be represented in a *confusion matrix*:

|               |       $Y = 1$        |       $Y = 0$        |
|---------------|:--------------------:|:--------------------:|
| $\hat{Y} = 1$ | True positives (TP)  | False positives (FP) |
| $\hat{Y} = 0$ | False negatives (FN) | True negatives (TN)  |

In a confusion matrix, correctly classified observations are on the
diagonal and off-diagonal values correspond to different types of
errors. Some of these error types are more relevant for certain
applications.

Imagine that you want to classify whether the water of a river is safe
to drink based on measurements of certain particles or chemicals in the
water (Y=1 means safe, Y=0 means unsafe). It's much worse to tag as
"safe" a polluted river than to tag as "unsafe" a potable water source,
one must be conservative. In this case, we would prioritize avoiding
false positives and wouldn't care so much about false negatives.

The following metrics are widely used and highlight different aspects of
our modeling goals.

-   **Accuracy** is simply the proportion of outputs that were correctly
    classified: $$ \text{Accuracy}=\frac{\text{TP} + \text{TN}}{N},$$
    where $N$ is the number of observations. This is a very common
    metric for training ML models and treats both classes as equally
    important. It's naturally extended to multi-class classification and
    usually compared to the value $\frac{1}{C}$ where $C$ is the number
    of classes.

Classification models are usually compared to randomness: How much
better is our model compared to throwing a coin for classification? At
random, we would assign each class $50\%$ of the time. So if we assume
that both classes are as likely to appear, that is, they are *balanced*,
the accuracy of a random guess would be around $0.5$. Hence, we want the
accuracy to be "better than random". If there are $C$ different classes
and the observations are balanced, we want the accuracy to be above
$1/C$.

A challenge is posed by *imbalanced* classes. For a dataset where $90\%$
of the observations are from class 1 and $10\%$ from class 0, always
predicting 1 would lead to a accuracy of $0.9$. This value may sound
good, but that model is not informative because it doesn't use any
information from predictors. Therefore, be careful when working with
imbalanced classes and interpreting your results.

-   **Precision** measures how often our "positive" predictions are
    correct:
    $$\text{precision} = \frac{\text{TP}}{\text{TP} + \text{FP}}.$$

-   The **true positive rate** (TPR), also called **Recall** or
    **sensitivity** measures the proportion of real "positives"
    ($Y = 1$) we are able to capture:
    $$ \text{TPR} = \frac{\text{TP}}{\text{TP}+\text{FN}}.$$

-   The **false positive rate** (FPR) is defined by
    $$\text{FPR} = \frac{\text{FP}}{\text{FP}+\text{TN}}.$$ and is
    related to another metric called **specificity** by
    $\text{FPR} = 1 - \text{specificity}$.

-   **Receiver operating characteristic (ROC) curve**: To evaluate the
    performance of a binary classification model, it's common to plot
    the *ROC curve*, where the TPR is plotted against the FPR, for
    varying values of the threshold $\tau$ used in the classification
    rule. When we decrease the threshold, we get more positive values
    (more observations are classified as 1), increasing both the true
    positive and false positive rate. The following image describes
    clearly how to interpret a ROC curve plot:

![ROC curves and how they compare, from
[Wikipedia](https://en.wikipedia.org/wiki/Receiver_operating_characteristic).](figures/Roc_curve.png)

<!-- XXX Exercises: Fit regression (NIGHT ~ SW), vary tau, plot TPR vs FPR. Fit second regression (NIGHT ~ TEMPERATURE), do the same. Interpret ROC curves. -->

-   **AUC**: The "area under the curve" is defined as the area left
    below the ROC curve. For a random classifier we would have AUC=0.5
    and for the perfect classifier, AUC=1. It's good to try to increase
    the AUC and it's used often as a reporting metric. Nevertheless, a
    visual inspection of the ROC curve can say even more.

-   **F1**: The F1 score is a more sophisticated metric, defined as the
    harmonic mean of precision and sensitivity, or in terms of the
    confusion matrix values: $$
    F1= 2 \times \frac{\text{precision} \times \text{recall}}{\text{precision} + \text{recall}} = \frac{2 \text{TP}}{2 \text{TP} + \text{FP} + \text{FN}}.
    $$ This metric provides good results for both balanced and
    imbalanced datasets and takes into account both the model's ability
    to capture positive cases (recall) and be correct with the cases it
    does capture (precision). It takes values between 0 and 1, with 1
    being the best and values of 0.5 and below being bad.

These metrics can be used to compare the quality of different
classifiers but also to understand the behaviour of a single classifier
from different perspectives.

This was an introduction of the most basic classification metrics. For a
more information on the topic, check out [this book
chapter](https://bookdown.org/max/FES/measuring-performance.html#class-metrics).

**Implementation in R**

Let's take a look at the previous metrics for the logistic regression
model we fitted before. The `confusionMatrix()` function from the
{caret} library provides most of the statistics introduced above.

```{r}
# Make classification predictions
Y <- logmod$data$Type
x <- as.factor(round(logmod$fitted.values)) # Use 0.5 as threshold

# Change class names
levels(Y) <- levels(x) <- c("Quebec", "Mississippi")

# plot confusion matrix
conf_matrix <- caret::confusionMatrix(data = x, reference = Y)
conf_matrix
```

Now we can visualize the confusion matrix as a mosaic plot. This is
quite helpful when we work with many classes.

```{r}
mosaicplot(conf_matrix$table,
           main = "Confusion matrix")
```

### Model evaluation

Model evaluation refers to several techniques that help you understand
how the model performs, whether this behavior is what you expect and how
you can improve it. You can use metrics and plots to get an overview of
the weaknesses of your model. This section covers model comparison,
variable selection and outlier detection, and more concepts related to
model evaluation (overfitting, data pre-processing, cross-validation...)
are explained in the remaining chapters. Concepts will be explained
using regression as an example, but are directly translated to
classification problems.

#### Model comparison

Be systematic with your model comparisons. Three key ideas in model
selection are:

-   Comparisons should be hierarchical: compare a model to another that
    "contains it", i.e. compare `GPP_NT_VUT_REF ~ SW_IN_F` to
    `GPP_NT_VUT_REF ~ SW_IN_F + LW_IN_F`, not `GPP_NT_VUT_REF ~ SW_IN_F`
    to `GPP_NT_VUT_REF ~ NIGHT + TA_F`.
-   Complexity must be increased slowly: add one variable at a time, not
    three variables all at once. This helps avoid collinearity in the
    predictors.
-   Choose the most appropriate metric: if possible, a metric that
    accounts for model complexity and represents the goal of your
    analysis (e.g., recall for a classification where you don't want to
    miss any positives).

If you're considering different model approaches for the same task, you
should first fit the best possible model for each approach, and then
compare those optimized models to each other. For example, fit the best
linear regression with your available data, the best KNN non-parametric
regression model and a random forest; then compare those three final
models and choose the one that answers your research question the best.

One must be careful not to keep training or improving models until they
fit the data perfectly, but maintain the models' ability to generalize
to newly available data. Chapter \@ref(supervisedmli) introduces the
concept of overfitting, which is central to data science. Think of model
interpretation and generalization when comparing them, not only of
performance. Simple models can be more valuable than very complex ones
because they tell a better story about the data (e.g., by having few
very good predictors rather than thousands of mediocre ones, from which
we cannot learn the underlying relationships).

#### Variable selection

Let's think of variable selection in the context of linear regression. A
brute force approach to variable selection would be: Fit a linear
regression for each combination of available predictors, calculate a
metric (e.g., AIC) and choose the best one (lowest AIC). The problem is,
if you have 8 predictors, you would fit 40320 different regression
models. This can be very computationally expensive.

Instead, take a hierarchical, or "greedy", approach, starting with an
empty model (just an intercept) and adding one variable at a time. This
is called **stepwise (forward) regression**. The algorithm goes as
follows:

-   First, you fit all regression models with just one variable and
    compute the $R^2$.
-   Then, select the one predictor leading to a model with the greatest
    $R^2$ (best fitting model) and compute the AIC (or BIC).
-   In the next step, compare remaining predictors to be added as a
    second variable in the model and calculating their $R^2$.
-   Choose as second predictor the one leading to the best $R^2$. Then,
    compute the AIC. If the AIC (which accounts for model fit and
    complexity) is worse, that is, bigger, stop and keep the univariate
    linear model. If the AIC is better, that is, smaller, add the second
    variable and repeat the previous steps to include a third variable.
    The method finishes once you cannot reduce the AIC anymore, or when
    you run out of variables. In the end, you'll have more or less the
    best possible linear regression model. The function `step()`
    implements the stepwise algorithm in R.

This stepwise approach can also be done backwards, starting with a full
model (all available variables) and removing one at a time. Or even with
a back-and-forth approach, where you look at both including a new or
removing an existing variable at each step (optimizing AIC).
Furthermore, this algorithm can be applied to fitting a polynomial
regression. We want to increase the degree of the polynomials unit by
unit. For a model with categorical variables, interaction terms should
only be considered after having the involved variables as "intercept
only".

**Multicollinearity** exists when there is a correlation between
multiple predictors in a multivariate regression model. This is
problematic because it makes the estimated coefficients corresponding to
the variables that are highly correlated very unstable. Since two highly
correlated variables explain almost the same, it doesn't matter whether
we include one or the other in the model (the performance metrics will
be similar) or even if we include both of them. Hence, it becomes
difficult to say which variables actually influence the target.

The **variance inflation factor (VIF)** is a score from economics that
measures the amount of multicollinearity in regression, based on how the
estimated variance of a coefficient is inflated due to its correlation
with another predictor. It's calculated as
$$\text{VIF}_i = \frac{1}{1 - R^2_i},$$ where $R^2_i$ is the coefficient
of determination for regressing the i$^{th}$ predictor on the remaining
ones. A VIF$_i$ is computed for each predictor in the multivariate
regression model we are evaluating, meaning: if $\text{VIF}_i = 1$
variables are not correlated; if $1 < \text{VIF}_i < 5$ there is
moderate collinearity; and if $\text{VIF}_i \geq 5$ they are highly
correlated. Because they can be almost fully explained by all the other
predictors (high $R^2_i$), these variables are redundant in our final
model.

To remedy collinearity, you may choose to use only one or two of those
correlated variables. Another option would be to use *Principal
Component Analysis* (PCA), which you may read more about
[here](https://www.r-bloggers.com/2021/05/principal-component-analysis-pca-in-r/).
What PCA does is to map the space of predictors into another of smaller
dimension, leading to a smaller set of predictor variables
$\{Z_1, ... , Z_q\}$, each of them being a linear combination of all the
initial available predictors, that is
$Z_1 = \alpha^1_0 + \alpha^1_1 X_1 + ... + \alpha^1_p X_p$, etc. If you
have collinearity, those highly correlated variables would be summarized
into one single new variable, called *principal component*.

When we work with high-dimensional data (that is, we have more variables
than observations) there are better techniques to do variable selection
than stepwise regression. Since the predictors space is so large, we
could fit a line that passes through all the observations (a perfect
fit), but does the model generalize? We don't know. For example, Lasso
and Ridge regression incorporate variable selection in the fitting
process (you can check [this
post](https://www.r-bloggers.com/2020/06/understanding-lasso-and-ridge-regression/)
if you're curious).

#### Outlier detection

Detecting outliers is important, because they can affect the fit of a
model and render the model fitting not robust. When the data is large,
individual points have less influence on the model fitting. Therefore
only outliers that are very far from normal values will affect the model
fit (see below). Outliers are particularly problematic in the context of
small data, because every value has a big influence on the fitted model.

Take a look at the two linear regressions below and how one single weird
observation can throw off the fit. Whenever an observation is very
distant from the center of the predictor's distribution, it becomes very
influential (it has a big *leverage*). If the observed response for that
data point is in harmony with the rest of points, nothing happens, but
if it's also off, the regression model will be affected greatly.

```{r warning=FALSE}
set.seed(2023)
hhdf_small <- hhdf |>
  sample_n(100) |> # reduce dataset
  select(SW_IN_F, GPP_NT_VUT_REF)

gg3 <- hhdf_small |>
  ggplot(aes(x = SW_IN_F, y = GPP_NT_VUT_REF)) +
  geom_point(size = 0.75) +
  geom_smooth(method = "lm", color = "red", fullrange = TRUE) +
  labs(x = expression(paste("Shortwave radiation (W m"^-2, ")")), 
       y = expression(paste("GPP (gC m"^-2, "s"^-1, ")"))) +
  theme_classic() +
  ylim(-20, 40) + 
  xlim(0, 2000)

gg4 <- hhdf_small |>
  add_row(SW_IN_F = 2000, GPP_NT_VUT_REF = -20) |> # add outlier
  ggplot(aes(x = SW_IN_F, y = GPP_NT_VUT_REF)) +
  geom_point(size = 0.75) +
  geom_smooth(method = "lm", color = "red", fullrange = TRUE) +
  labs(x = expression(paste("Shortwave radiation (W m"^-2, ")")), 
       y = expression(paste("GPP (gC m"^-2, "s"^-1, ")"))) +
  theme_classic() +
  geom_point(aes(x = 2000, y = -20), colour='blue') +
  ylim(-20, 40) + 
  xlim(0, 2000)
  
cowplot::plot_grid(gg3, gg4)
```

The first step to identifying outliers is to look at your data, one
variable at a time. Plot a histogram to see the rough distribution of a
variable. This will help identify what kind of values to expect. In
Chapters \@ref(datawrangling) and \@ref(datavis) it was introduced how
to identify values that fell out of this distribution using histograms
and boxplots. Checking in the histogram if the distribution has fat
tails helps to discern whether the values that pop out of a boxplot
should be considered outliers or not.

```{r}
gg5 <- hhdf_small |>
  add_row(SW_IN_F = 2000, GPP_NT_VUT_REF = -20) |> # add outlier
  ggplot(aes(x = GPP_NT_VUT_REF, y = after_stat(density))) +
  geom_histogram(fill = "grey70", color = "black") +
  geom_density(color = 'red')+
  labs(title = 'Histogram, density and boxplot', 
       x = expression(paste("GPP (gC m"^-2, "s"^-1, ")"))) +
  theme_classic()

gg6 <- hhdf_small |>
  add_row(SW_IN_F = 2000, GPP_NT_VUT_REF = -20) |> # add outlier
  ggplot(aes(x = "", y = GPP_NT_VUT_REF)) +
  geom_boxplot(fill = "grey70", color = "black") +
  coord_flip() +
  theme_classic() +
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
  labs(y = expression(paste("GPP (gC m"^-2, "s"^-1, ")")))

gg7 <- hhdf_small |>
  add_row(SW_IN_F = 2000, GPP_NT_VUT_REF = -20) |> # add outlier
  ggplot(aes(x = SW_IN_F, y = after_stat(density))) +
  geom_histogram(fill = "grey70", color = "black") +
  geom_density(color = 'red')+
  labs(title = 'Histogram, density and boxplot', 
       x = expression(paste("Shortwave radiation (W m"^-2, ")"))) +
  theme_classic()

gg8 <- hhdf_small |>
  add_row(SW_IN_F = 2000, GPP_NT_VUT_REF = -20) |> # add outlier
  ggplot(aes(x = "", y = SW_IN_F)) +
  geom_boxplot(fill = "grey70", color = "black") +
  coord_flip() +
  theme_classic() +
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
  labs(y = expression(paste("Shortwave radiation (W m"^-2, ")")))

cowplot::plot_grid(gg5, gg7, gg6, gg8,
                   ncol = 2, rel_heights = c(2,1),
                   align = 'v', axis = 'lr')
```

A **Q-Q Plot** depicts the sample quantiles of a variable against the
theoretical quantiles of a distribution of our choice, usually a normal
distribution. In the histograms above, GPP looks somewhat Gaussian but
with fatter tails and slightly skewed to the right, while shortwave
radiation is skewed to the right, resembling an exponential
distribution. This is also visible in the Q-Q plots below, because
outliers deviate greatly from the straight line (which represents a
match between the observed values and the theoretical distribution):

```{r}
gg9 <- hhdf_small |>
  add_row(SW_IN_F = 2000, GPP_NT_VUT_REF = -20) |> # add outlier |>
  ggplot(aes(sample = GPP_NT_VUT_REF)) +
  geom_qq() +
  geom_qq_line() +
  labs(y = expression(paste("GPP (gC m"^-2, "s"^-1, ")")),
       x = "Theoretical normal quantiles") +
  theme_classic()

gg10 <- hhdf_small |>
  add_row(SW_IN_F = 2000, GPP_NT_VUT_REF = -20) |> # add outlier |>
  ggplot(aes(sample = SW_IN_F)) +
  geom_qq() +
  geom_qq_line() +
  labs(y = expression(paste("Shortwave radiation (W m"^-2, ")")),
       x = "Theoretical normal quantiles") +
  theme_classic()

cowplot::plot_grid(gg9, gg10, ncol=2)
```

For linear (and logistic) regression, we would like predictor variables
to look as normal, i.e. gaussian, as possible. You've probably learned some of the
reasons for this in quantitative methods courses, but are beyond the
scope of this class. It's common to study the distribution of the
regression residuals with QQ-plots to assess if model assumptions are
met.

Above, you can see the distributions of our target and predictor (with
outliers). And it's very easy to see the weird value for the shortwave
radiation but for GPP it doesn't stick out so much. This already points
to how important it is to check their multivariate distribution. R
provides some useful plots from the fitted regression objects, in
particular the "Residuals vs Leverage" plot:

```{r}
# Fit regression with outlier
linmod_outlier <- lm(GPP_NT_VUT_REF ~ SW_IN_F, 
                     data = add_row(hhdf_small, SW_IN_F = 2000, GPP_NT_VUT_REF = -20))

plot(linmod_outlier, 5)
```

This plot shows the leverage (see the mathematical definition
[here](https://en.wikipedia.org/wiki/Leverage_(statistics))) of each
observation against the corresponding residual from the fitted linear
regression. Points with high leverage, i.e., points that are far from
the center of the predictor distribution, and large residuals, i.e.,
points that are far from the fitted regression line, are very
influential. **Cook's distance** (definition
[here](https://en.wikipedia.org/wiki/Cook%27s_distance)) is an estimate
of the influence of a data point in a linear regression and observations
with Cook's distance \> 1 are candidates for being outliers. See in the
plot above how the point with index 101 (our added outlier) has a very
large Cook's distance. Boundary regions for Cook's distance equal to 0.5
(suspicious) and 1 (certainly influential) are drawn with a dashed line.

Finally, it's very important that, before you remove a value because it
may be an outlier, you understand where the data came from and if such
an abnormal observation is possible. If it depicts an extraordinary but
possible situation, this information can be very valuable and it's wiser
to keep it in the model. Interesting research questions arise when data
doesn't align with our preconceptions, so keep looking into it and
potentially collect more data.

## Report Exercise

<!-- -   Implement the formula for RMSE using simple "low-level" functions -->
<!--     like `sqrt()` and `mean()`. Confirm that the function `rmse()` from -->
<!--     the yardstick package computes the RMSE the same way. -->
<!--     ```{r} -->
<!--     ## generate random data  -->
<!--     df <- tibble(x = rnorm(100)) %>%  -->
<!--       mutate(y = x + rnorm(100), -->
<!--              x = x) -->
<!--     yardstick::rmse(df, y, x) %>% pull(.estimate) -->
<!--     sqrt(mean((df$x - df$y)^2)) -->
<!--     ``` -->

This tutorial's exercise is a bit special. Although there is a lot of helpful information out there, you will often have to write your own data analysis routine. This requires good understanding of statistical knowledge, algorithmic thinking and problem-solving skills. While writing your code, you will face many questions and bugs that you need to solve. And knowing where and how to ask for help properly are crucial parts of this process (see second chapter for more on getting help). To learn these skills, you will attempt to write your own step-wise forward regression from scratch using the theory and code provided this and previous chapters (see task statement and hints below).

**Important:** We do not expect that all of you solve this exercise without any issues - which is completely fine! **Thus, in your final report**, you will have to add a minimum reproducible example (MRE), following [this guideline](%5Bgreat%20template%5D(https://stackoverflow.com/questions/5963269/how-to-make-a-great-r-reproducible-example)). Write this MRE as if you were addressing an online audience; describe in detail what you goal is, where you got stuck, what the error message is, provide a code example that is runnable without needing the .csv files that you have locally (see `dput()`), etc. If you can solve this problem without any issues, congratulations! But it is likely that you did not do so without facing any error messages - take one of these errors and write a MRE.

Important tools to code a step-wise regression are nested for-loops, if-statements, and finding the best predictor for a given target variable. Instead of trying all of this at once, it is advisable to test your code in separate pieces that can then be stitched together. To help you out with these concepts, we provide two "warm-up exercises" (solutions are given at the end of this book).

### Task Statement

Re-read the section on stepwise regression of this tutorial and write a pseudo-code for how you would encode a stepwise regression, write-out using coding terms like "I have to loop over all predictors and for each predictor I have to check...". For picking the next-best variable to add, you can pick your metric of choice: $R^2$, *adjusted-*$R^2$, the Akaike Information Criterion (AIC) or the Bayesian Information Criterion (BIC). Visualise the result of the step-wise regression so that it shows how the chosen metric varies with the number of predictors, preferably showing the name of the additional predictor (not that you don't always have to use graphs, you can also create a simple table for this). Use the code chunk below to start with an already cleaned dataset.



```{r}
df <- read_csv("df_for_stepwise_regression.csv")
```

*Hints*:

-   Really try to think of the blueprint (*pseudo-code*) first: How to go through different models in each forward step? How to store predictors added to the model and how to update candidate predictors?

-   You will have to add predictors sequentally to your final model as you search through all predictors. You can use `new_list <- list()` to create an empty vector, and then add elements to that list by `new_list <- c(new_list, new_element)`.

-   It may be helpful to explicitly define a set of "candidate predictors" that are potentially added to the model (e.g., `preds_candidate`), and define predictors retained in the model from the previous step in a separate vector (e.g., `preds_retained`). In each step, search through `preds_candidate`, select the best predictor, add it to `preds_retained` and remove it from `preds_candidate`.

-   At each step, record the metrics and store them in a data frame for later plots. As in the first "warm-up" exercise, you may record metrics at each step as a single-row data frame and sequentially stack (bind) them together.

-   You will have to create new formulas for models dynamically, based on the candidate predictors. A clever way how to construct formulas dynamically is described in [this stackoverflow post](https://stackoverflow.com/questions/4951442/formula-with-dynamic-number-of-variables).

<!-- -   The metrics for each model are assessed *after* the order of added variables is determined. To be able to determine the metrics, the models can be saved by constructing a list of models and sequentially add elements to that list (`mylist[[ name_new_element ]] <- new_element`). You can also fit the model again after determining which predictor worked best. -->

<!-- -   Your code will most certainly have bugs at first. To debug efficiently, write code first in a simple R script and use the debugging options in RStudio (see [here](https://support.rstudio.com/hc/en-us/articles/205612627-Debugging-with-RStudio)). -->


-   For visualising: To display a table nicely as part of the RMarkdown html output, use the function `knitr::kable()`. To avoid reordering the list of variable names when plotting, change the type of variable names from "character" to "factor" by `pred <- factor(pred, levels = pred)`

### Warm-Up Exercises

#### Nested loops and if-statements

Given a matrix A and a vector B (see below), do the following tasks:

-   Replace the missing values (`NA`) in the first row of A by the largest value of B. After using that element of B for imputing A, drop that element from the vector B and proceed with imputing the second row of A, using the (now) largest value of the updated vector B, and drop that element from B after using it for imputing A. Repeat the same procedure for all four rows in A.

-   After imputing (replacing) in each step, calculate the mean of the remaining values in B and record it as a single-row data frame with two columns `row_number` and `avg`, where `row_number` is the row number of A where the value was imputed, and `avg` is the mean of remaining values in B. As the algorithm proceeds through rows in A, sequentially bind the single-row data frame together so that after completion of the algorithm, the data frame contains four rows (corresponding to the number of rows in A).

```{r}
A <- matrix(c(6, 7, 3, NA, 15, 6, 7, 
              8, 9, 12, 6, 11, NA, 3, 
              9, 4, 7, 3, 21, NA, 6, 
              7, 19, 6, NA, 15, 8, 10),
            nrow = 4, byrow = TRUE)
B <- c(8, 4, 12, 9, 15, 6)
```

Before implementing these tasks, try to write down a pseudo code. This is code-like text that may not be executable, but describes the structure of real code and details where and how major steps are implemented. Next, you'll need to write actual R code. For this, you will need to find answers to the following questions:

-   How to go through each of the element in matrix?
-   How to detect NA value?
-   How to drop an element of a given value from a vector?
-   How to add a row to an existing data frame?

#### Find the single best predictor

The first step of a stepwise forward regression is to find the single most powerful predictor in a univariate linear regression model for the target variable `GPP_NT_VUT_REF` among all available predictors in our data set (all except those of type `date` or `character`). Implement this first part of the search, using the definition of the stepwise-forward algorithm below. Remove all rows with at least one missing value before starting the predictor search.

*Algorithm:*

1.  Let $\mathcal{M_0}$ denote the null model, which contains no predictors.

2.  For $k=0,..,p-1$:
    (a) Consider all $p  k$ models that augment $\mathcal{M}_k$ with one additional predictor.
    (b) Choose the best model among these $p  k$ models, and call it $\mathcal{M}_{k+1}$. Here *best* is defined as having the highest $R^2$ .

*Hints*:

-   The "counter" variables in the for loop can be provided as a vector, and the counter will sequentially take on the value of each element in that vector. For example:

```{r}
all_predictors <- df |>
  select(-GPP_NT_VUT_REF) |>
  names()

cat("Predictor variables and their first 3 entries in df: \n")

for (var in all_predictors){
  cat("\n", var, ": ")
  cat(head(df[[var]], 3))
  # cat("\n")
}
```

-   To record $R^2$ values for the different models, you may start by creating an empty vector (`vec <- c()`) before the loop and then sequentially add elements to that vector inside the loop (`vec <- c(vec, new_element)`). Alternatively, you can do something similar, but with a data frame (initialising with `df_rsq <- data.frame()` before the loop, and adding rows by `df_rsq <- bind_rows(df_rsq, data.frame(pred = predictor_name, rsq = rsq_result))` inside the loop).
