# (APPENDIX) Appendix {-}

# Solutions {#solutions}

## Getting Started

### Dimensions of a circle {-}

- Given the radius of a circle write a few lines of code that calculates its area and its circumference. Run your code with different values assigned of the radius. 
```{r}
radius <- 1
area <- pi * radius^2
circum <- 2 * pi * radius
```
- Print the solution as text.
```{r}
print(paste("Radius:", radius, "   Circumference:", circum))
```

### Sequence of numbers {-}

Generate a sequence of numbers from 0 and $\pi$ as a vector with length 5.
```{r}
seq(0, pi, length.out = 5)
```

### Gauss sum {-}

Rumors have it that young Carl Friedrich Gauss was asked in primary school to calculate the sum of all natural numbers between 1 and 100. He did it in his head in no time. We're very likely not as intelligent as young Gauss. But we have R. What's the solution?

```{r}
sum(1:100)
```

### Magic trick algorithm {-}

Define a variable named `x` that contains an integer value and perform the following operations in sequence:

- Redefine `x` by adding 1.
- Double the resulting number, over-writing `x`.
- Add 4 to `x` and save the result as `x`.
- Redefine `x` as half of the previous value of `x`.
- Subtract the originally chosen arbitrary number from `x`.

Print `x`. Restart the algorithm defined above by choosing a new arbitrary natural number between 1 and 20.

```{r}
x <- -999  # arbitrary number between 1 and 20
x_save <- x  # save for the last step
x <- x + 1
x <- x * 2
x <- x + 4
x <- x / 2
x - x_save
```

### Vectors {-}

Print the object `datasets::rivers` and consult the manual of this object.

- What is the class of the object? 
- What is the length of the object?
- Calculate the mean, median, minimum, maximum, and the 33%-quantile across all values. 

```{r}
class(datasets::rivers)
length(datasets::rivers)
mean(datasets::rivers)
# other functions can easily be found on the internet ;-)
```

### Data frames {-}

Print the object `datasets::quakes` and consult the manual of this object.

- Determine the dimensions of the data frame using the respective function in R.
- Extract the vector of values in the data frame that contain information about the Richter Magnitude.
- Determine the value largest value in the vector of event magnitudes.
- Determine the geographic position of the epicenter of the largest event.

```{r}
dim(datasets::quakes)
vec <- datasets::quakes$mag
max(vec)
idx <- which.max(vec)  # index of largest value

# geographic positions defined by longitude and latitude (columns long and lat)
datasets::quakes$long[idx]
datasets::quakes$lat[idx]
```

### RMarkdown {-}

Create an RMarkdown file and implement your solutions to above exercises in it. Give the file a title, implement some structure in the document, and write some text explaining what your code does.

